---
title: "Project Assignment 1"
author: "Mark Sandstrom"
date: '2016-01-09'
output:
  html_document:
    keep_md: yes
keep_md: yes
---
Reproducing results on analyzing steps taken by a volunteer at different times of day
---
**1. Code for reading in the dataset and/or processing the data**

To reproduce our results, first copy and unzip the measurement data file activity.csv from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip to the R working directory. Next, read it into a data frame, and omit the rows with no values in in the measurement column `steps`:
```{r, echo=T, meassage=F, warning=F}
library(plyr)
dfwna=read.csv('activity.csv')
df=subset(dfwna,steps!='NA')
any(is.na(df))
```

To view basic aspects of the measurement data frame `df`:
```{r, echo=T, meassage=F, warning=F}
str(df)
summary(df)
```

The measurement `steps` is simply an integer count of (e.g. walking) steps taken by the volunteer wearing the monitoring system on the `date` and the observation `interval` by their respective columns, where the value of `interval` is an identification of a 5-minute interval during any given `date`.

**2. Histogram of the total number of steps taken each day**

To histogram the daily totals of the `steps` with measurement intervals with missing values (NAs) for `steps` ignored:
```{r, echo=T, meassage=F, warning=F}
dailytotals=ddply(df, c('date'), summarise, totals=sum(steps))
hist(dailytotals$totals, 100, col='green', xlab='daily-totals steps (ingoring NAs)', ylab='count of days with the given daily-total')
```

**3. Mean and median number of steps taken each day**

To report the daily means and medians of `steps` taken:
```{r, echo=T, meassage=F, warning=F}
summary(round(dailytotals$totals))
```

**4. Time series plot of the average number of steps taken**

To see any pattern in the daily averages of `steps`:
```{r, echo=T, meassage=F, warning=F}
plot(dailyaves$date, dailyaves$mean, xlab='date', ylab='daily averages of steps for 5-minute intervals (omitting NAs)')
lines(dailyaves$date, dailyaves$mean)
```

**5. The 5-minute interval that, on average, contains the maximum number of steps**

To see any pattern in the per-5-minute-interval averages of `steps` across the dates:
```{r, echo=T, meassage=F, warning=F}
intervalaves=ddply(df, c('interval'), summarise, mean=mean(steps))
plot(intervalaves$interval, intervalaves$mean, type='l', col='red', xlab='interval', ylab='mean steps')
```

The 5-minute `interval` with maximum `steps` in average across the `dates`:
```{r, echo=T, meassage=F, warning=F}
maxi=subset(intervalaves, mean==max(intervalaves$mean))
maxi
```

**6. Code to describe and show a strategy for imputing missing data**

To replase the missing values for `steps` by the average value for their respective intervals:
```{r, echo=T, meassage=F, warning=F}
dfim=dfwna
dfim$steps[is.na(dfim$steps)] = with(dfim, ave(steps, interval, FUN = function(x) mean(x, na.rm=T)))[is.na(dfim$steps)]
any(is.na(dfim))
```

**7. Histogram of the total number of steps taken each day after missing values are imputed**

To histogram the daily totals of the `steps` with measurement intervals after missing values (NAs) for `steps` imputed per above:
```{r, echo=T, meassage=F, warning=F}
dailytotalsim=ddply(dfim, c('date'), summarise, totals=sum(steps))
hist(dailytotalsim$totals, 100, col='blue', xlab='daily-total steps (NAs replaced by interval-means)', ylab='count of days with the given daily-total')
```

**8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends**

To see if there is a difference in per-interval-averages of steps taken between weekdays (Mon-Fri) and weekends (Sat-Sun):
```{r, echo=T, meassage=F, warning=F}
dfwd=mutate(df, wd=timeDate::isWeekday(date))
dfwk=subset(dfwd,wd==T)
dfwe=subset(dfwd,wd==F)
par(mfrow=c(2,1))
intervalaves=ddply(dfwk, c('interval'), summarise, mean=mean(steps))
plot(intervalaves$interval, intervalaves$mean, 'l', col='blue', xlab='interval', ylab='Mon-Fri mean steps',lwd=3)
intervalaves=ddply(dfwe, c('interval'), summarise, mean=mean(steps))
plot(intervalaves$interval, intervalaves$mean, 'l', col='green', xlab='interval', ylab='Sat-Sun mean steps',lwd=3)

```

